SAT Solving for Vesicle Traffic Systems in Cells

### 2017-06-28 paper/intro.tex
fixed_bib

The conservation of basic molecular processes across all living cells, including the genetic code, suggests that life on Earth arose only once: all extant organisms are presumed to trace their lineages back to a last universal common ancestor (LUCA). However, there are still vast differences in cellular and molecular architecture among the three kingdoms of life: bacteria, archaea, and eukaryotes [1]. Eukaryotic cells have larger genomes and have more complex cellular architectures than the prokaryotic bacteria and archaea. In particular, the cytoplasm of all eukaryotic cells is broken up into distinct membrane-bound locations known as organelles: the nuclear membrane, endoplasmic reticulum (ER), Golgi apparatus, vacuoles and lysosomes are examples of these. Cargo is moved between these organelles in small membrane-bound transporters known as vesicles [2].

This organization resembles a transport logistics network, and is commonly known as the membrane traffic system. Membrane traffic underpins nearly every aspect of eukaryotic cellular physiology, including human cellular physiology. Defects in membrane traffic lead to a variety of disorders or even cell death [2]. Understanding how membrane traffic functions is therefore one of the central challenges of cell biology. Some progress has been made, as cell biologists have assembled a “parts list” of molecules that drive vesicle traffic [3]. The essential processes involve the creation of vesicles loaded with cargo from source organelles, known as “budding”, and the depositing of those vesicles and cargo into target organelles, known as “fusion” [4]. Budding is regulated by proteins known as coats and adaptors that select cargo. Fusion is regulated by proteins known as SNAREs and tethers, that ensure that vesicles fuse to the correct target [5]. Although we know a great deal about molecular-level details, there has been very little work done on how such processes are integrated across scales to build the entire vesicle traffic system. We have recently attempted to use ideas of graph theory to address this question [5,6]. The SNARE proteins are the focus of our current work. Abstractly, we can think of these proteins as labels that are collected from the source compartment and taken to the target compartment on vesicles. There is a corresponding set of label molecules on the target compartments. If the two sets of labels (on vesicles and targets) are compatible, then the vesicle will fuse. The question is: does this physical picture place any constrains on the global topology of the traffic graph? We have previously shown that one informative constraint is graph connectivity [6].

1. Embley, T. M., and Martin, W. (2006). Eukaryotic evolution: changes and challenges. Nature 440, 623-630.

2. Stenmark, Harald. "Rab GTPases as coordinators of vesicle traffic." Nature Reviews Molecular Cell Biology 10, no. 8 (2009): 513-525.

3. Dacks, J. B., & Field, M. C. (2007). Evolution of the eukaryotic membrane-trafficking system: origin, tempo and mode. Journal of cell science, 120(17), 2977-2985.

4. Munro S. 2004. Organelle identity and the organization of membrane traffic. Nature Cell Biology 6:469–472. doi: 10.1038/ncb0604-469.

5. Mani, S., and Thattai, M. (2016b). Stacking the odds for Golgi cisternal maturation. eLife 10.7554/eLife.16231.

6. Shukla, A., Bhattacharyya, A., Kuppusamy, L., Srivas, M., & Thattai, M. (2017). Discovering vesicle traffic network constraints by model checking. PLoS ONE, in press.

@@

vesicle traffic systems(VTS)

>>>>

The conservation of basic molecular processes across all living cells, including the genetic code, suggests that life on Earth arose only once: all extant organisms are presumed to trace their lineages back to a last universal common ancestor (LUCA). However, there are still vast differences in cellular and molecular architecture among the three kingdoms of life: bacteria, archaea, and eukaryotes \cite{embley2006eukaryotic}. Eukaryotic cells have larger genomes and have more complex cellular architectures than the prokaryotic bacteria and archaea. In particular, the cytoplasm of all eukaryotic cells is broken up into distinct membrane-bound locations known as organelles: the nuclear membrane, endoplasmic reticulum (ER), Golgi apparatus, vacuoles and lysosomes are examples of these. Cargo is moved between these organelles in small membrane-bound transporters known as vesicles \cite{stenmark2009rab}.

This organization resembles a transport logistics network, and is commonly known as the membrane traffic system. Membrane traffic underpins nearly every aspect of eukaryotic cellular physiology, including human cellular physiology. Defects in membrane traffic lead to a variety of disorders or even cell death \cite{stenmark2009rab}. Understanding how membrane traffic functions is therefore one of the central challenges of cell biology. Some progress has been made, as cell biologists have assembled a “parts list” of molecules that drive vesicle traffic \cite{dacks2007evolution}. The essential processes involve the creation of vesicles loaded with cargo from source organelles, known as “budding”, and the depositing of those vesicles and cargo into target organelles, known as “fusion” \cite{munro2004organelle}. Budding is regulated by proteins known as coats and adaptors that select cargo. Fusion is regulated by proteins known as SNAREs and tethers, that ensure that vesicles fuse to the correct target \cite{mani2016stacking}. Although we know a great deal about molecular-level details, there has been very little work done on how such processes are integrated across scales to build the entire vesicle traffic system. We have recently attempted to use ideas of graph theory to address this question \cite{mani2016stacking,shukla}. The SNARE proteins are the focus of our current work. Abstractly, we can think of these proteins as labels that are collected from the source compartment and taken to the target compartment on vesicles. There is a corresponding set of label molecules on the target compartments. If the two sets of labels (on vesicles and targets) are compatible, then the vesicle will fuse. The question is: does this physical picture place any constrains on the global topology of the traffic graph? We have previously shown that one informative constraint is graph connectivity \cite{shukla}.

@@

Vesicle traffic systems(VTS)

### 2017-06-28 paper/related.tex
fixed_bib

Satisfiability was the first NP-complete problem discovered by Cook in his famous paper [ref]. Every NP-complete problem can be reduced to solving for satisfiability. The fact is that SAT solvers have seen a surge in the handling of millions of variables, hence many general problems are encoded as a SAT problem. Due to the nature of the searching (exhaustive), solving the combinatorial problem is a natural fit for the SAT solvers. There are many graphs related combinatorial work [1-4] that have used SAT solvers. Thanks to the improvement in SAT solver and formal techniques, tools like model checkers and theorem provers scalability and combinatorial solving capability are now being used to model complex biological problems [1-3].
% Many network questions are being reduced to solving a SAT question.
Many of these biological system uses model checkers to reason about graphs and networks rules with possible exhaustive search [5-6].

Recently model checkers found the use of modeling and understanding the gene regulatory networks (GNR) [7-9]. GNR is a complex system driven by many complex rules. We extended this work and have applied model checkers to the more complex transport network (VTS) [Mani and Thattai]. Model checkers are far scalable in comparison to simulations, but scaling it to cases which are interesting biologically is still a challenge. In this paper we have achieved the scalability required for analyzing vesicular transport network for eukaryotic cells, of total ten compartments (N = 10).

\textbf{SAT, Bio and Graph}
\begin{enumerate}
\item Reasoning over Biological Networks using Maximum Satisfiability.
\item BioGraphE: high-performance bionetwork analysis using the Biological Graph Environment.
\item Generalized k-ary tanglegrams on level graphs: A satisfiability-based approach and its evaluation.
\item SMT-Based Analysis of Biological Computation.
\item A SAT-based algorithm for finding attractors in synchronous Boolean networks.
\item ILP/SMT-Based Method for Design of Boolean Networks Based on Singleton Attractors.
\item Exact DFA Identification Using SAT Solvers
\item Timing Robustness in the Budding and Fission Yeast Cell Cycles.
\item Solving Subgraph Epimorphism Problems using CLP
and SAT.
\end{enumerate}

\textbf{GNR and SAT}
\begin{enumerate}
\item Model Checking Gene Regulatory Networks.
\item Inference of Boolean Networks from Gene Interaction Graphs Using a SAT Solver
\item Efficient parameter search for qualitative models of regulatory networks using symbolic model checking.
\end{enumerate}

>>>>

Satisfiability was the first NP-complete problem discovered by Cook in his famous paper [ref]. Every NP-complete problem can be reduced to solving for satisfiability. The fact is that SAT solvers have seen a surge in the handling of millions of variables, hence many general problems are encoded as a SAT problem. Due to the nature of the searching (exhaustive), solving the combinatorial problem is a natural fit for the SAT solvers. There are many graphs related combinatorial work \cite{gay2013solving,wotzlaw2012generalized} that have used SAT solvers. Thanks to the improvement in SAT solver and formal techniques, tools like model checkers and theorem provers scalability and combinatorial solving capability are now being used to model complex biological systems \cite{heule2010exact,yordanov2013smt,mangla2010timing}.
% Many network questions are being reduced to solving a SAT question.
Many of these biological system uses model checkers to reason about graphs and networks rules with possible exhaustive search \cite{guerra2012reasoning,chin2008biographe}.

Recently model checkers found the use of modeling and understanding the gene regulatory networks (GNR) \cite{giacobbe2015model,rosenblueth2014inference, batt2010efficient}. GNR is a complex system driven by many complex rules. We extended this work and have applied model checkers to the more complex transport network (VTS) \cite{mani2016stacking}. Model checkers are far scalable in comparison to simulations, but scaling it to cases which are interesting biologically is still a challenge. In this paper we have achieved the scalability required for analyzing vesicular transport network for eukaryotic cells, of total ten compartments (N = 10). \\
